---
title: "Reproducible Research: Peer Assessment 1"
author: "Carolina San Martín"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Charging required libraries
library(ggplot2)
library(dplyr)

```

```{r load and clean data, message=FALSE, warning=FALSE}
# Unzip and read csv data
activity <- unzip("activity.zip")
raw_data <- read.csv("activity.csv")
rm(activity)

# Primary clean-up: omit NA values and convert data variable to date format
data <- na.omit(raw_data)
date <- as.Date(data$date)
```


## What is mean total number of steps taken per day?
```{r Descriptive analysis, fig.height=4, message=FALSE, warning=FALSE, results="hide"}
# Daily steps
steps_day <- data %>% 
    group_by(date) %>% 
    summarise(total_steps=sum(steps))

# Graph daily data in an histogram
hist(x = steps_day$total_steps, 
     main = "Histogram of number of steps each day", 
     xlab = "Total steps each day", ylab = "Frequency",
     col = "purple")

# Descriptive statistics
mean_steps <- mean(steps_day$total_steps)
median_steps <- median(steps_day$total_steps)
```

The mean of total number of steps taken per day is `r mean_steps` and the median of total steps by day is `r median_steps`.

## What is the average daily activity pattern?
```{r daily-pattern, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
#Data by 5 minutes-interval
interval <- as.numeric(data$interval)
five_av <- data %>% 
    group_by(interval) %>% 
    summarise(av_steps=mean(steps))

# Average activity time-series plot
plot(five_av$interval, five_av$av_steps, 
     main="Average steps taken in a 5-minute interval",
     type="l",
     xlab="Interval", ylab="Average steps",
     col = "purple")

# Interval with maximum number of steps
max <- five_av$interval[which.max(five_av$av_steps)]
```

The interval with the highest steps counted happened during `r max`.


## Imputing missing values
```{r na-values}
# Calculate number of missing values
na <- sum(is.na(raw_data$steps))
```

There are `r na` data with NA values in the raw data.

```{r filling na-values, message=FALSE, warning=FALSE}
# Create filled dataset assigning mean value to steps variable. First, find the value when interval matches the average, and then, assign the value instead the NA.
fill_data <- raw_data
for (j in 1:nrow(fill_data)) {
    if (is.na(fill_data$steps[j])) {
        replacing <- which(fill_data$interval[j]==five_av$interval)
        fill_data$steps[j] <- five_av[replacing,]$av_steps
    }
}

# Create data by date using prior filled dataset
fill_stepsd <- fill_data %>% 
    group_by(date) %>% 
    summarise(total_filled_steps=sum(steps))

# Graph with filled data by day
hist(x = fill_stepsd$total_filled_steps, 
     main = "Histogram of total number of steps each day with filled dataset", 
     xlab = "Total steps each day", ylab = "Frequency",
     col = "purple")

# Descriptive statistics
filled_mean <- mean(fill_stepsd$total_filled_steps)
filled_median <- median(fill_stepsd$total_filled_steps)
```

The mean and median total number of steps taken per day are `r filled_mean` and `r filled_median`, respectively. The mean is unchanged from the estimates from the first part of the assignment, but the median changed slightly and now is equal to the mean. Imputing missing data based on an average value only affected to median value. 

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays and weekend analysis, message=FALSE, warning=FALSE}
# Create weekday variable using weekdays()
fill_data$date <- as.Date(fill_data$date)
fill_data$day <- weekdays(fill_data$date)

# Define all days as weekdays
fill_data$day_type <- "weekday"

# Fix days that are from weekend
fill_data$day_type[fill_data$day %in% c("Saturday", "Sunday")] <- "weekend"

# Create data by 5-minute interval with filled data
day_av <- fill_data %>% 
    group_by(day_type, interval) %>% 
    summarise(average_steps=mean(steps))

# Graph the two-in-one time-series graphic.
qplot(interval, average_steps, data=day_av,
      main="Average steps comparing Weekends vs. Weekdays",
      type="l",
      geom="line",
      xlab="Interval", ylab="Average of Steps",
      facets =day_type ~ .)
```

Plot presents differences between weekdays and weekend: there are more activity measured in steps on weekdays.

